### 2024-06-01 README.Rmd
I am now polishing the first part, and tying up the second.

Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_. _No generative AI tool was used to research or write this essay_.

@@

- [Background](#background)  
- [The university's mission](#the-mission)  
- [What do AIs do?](#what-do-ais-do)  
- [A note on the terms of the conversation](#terms-of-debate)  
- [But really, what do AIs do?](#really-what-do-they-do)  
- [AI in teaching and learning](#ai-in-teaching-and-learning)  
- [What can GenAI tools do for students?](#what-can-ais-do-for-students)  
- [What do GenAI tools do to students?](#what-can-ais-do-to-students)  
- [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
- [What do GenAIs to to everyone else?](#what-do-ais-do-to-everyone-else)

@@

GenAI was announced as nothing short of revolutionary. McMaster, for example, [stated](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:

@@

AI was quickly painted as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding other things for people to do.]. It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.

The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students], has led to institutional responses that have been less than coherent. A case in point is Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].

@@

The sub-committees, in turn, convened panels to [provide recommendations]({#ai-panel-tasks}) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/). Participants in the panel were told that they would help to shape the guidelines for teaching and learning at McMaster, a consequential task "for how educators take up generative AI in their course and assessment design and how students approach the use of this technology"^[I have been unable to find the terms of reference of the panels. Who will decide which recommendations of the panel are adopted and which are ignored?].

@@

Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption. This attitude is almost fatalistic^[In the sense of "fated to"] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one with chatbots in the pole position is just the most recent of several hype cycles [see @siegel_ai_2023]] [@angwin_press_2024].

@@

(The last sentence in the statement above carries quite a bit of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.)

To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AIs, it is essential to have a working understanding of what the technology claims to do that sets it apart from other tools.

@@

Clearly, there is some tension implicit in the mission of a university. But being generous, we can take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with integrity and quality.

@@

A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they are seen as _everything tools_. While hammers are understood to have a limited range of uses having to do with concentrating force, and word processors are mostly used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].

The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023]; that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023]; according to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024]. But there is no reason to stop at self-driving vehicles: McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023]. In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].

AI would even write code, perhaps helping itself to some self-programming along the way: AIs that write themselves, or AI self-help?

Figure @\ref(fig:ai-writing-code)

```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure X. An example of Atlassian AI writing code"}
knitr::include_graphics("images/AI-bug-fix.png")
```

The list of promises about what AI will do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].

Higher-education has not spared the onslaught of promises.

@power_3_2024,^[Rhett Power is a motivational speaker and executive coach] for example, presents three innovative use cases for AI in higher education. First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies"]. AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants. AI could also personalize the experience of consumers (also called students) by predicting their final grade with 60-70% accuracy, and by providing corrective and instructive feedback.
<!--release them in a very literal sense, as one advantage is to do _more_ with limited resources (which in the current fiscal environment means laying people off). The cases discussed by Mr. Power are questionable: there are already non-AI systems that deploy paperwork; the kind of information needed to identify students at risk of dropping out is probably quite invasive, and therefore impractical or unethical to collect; I very much doubt that it is legal to use AI in hiring and admissions. --> Mr. Power's message to the higher-education sector? It is time for institutions to "level up": jump on the bandwagon, fellas, don't let it let you behind!

In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school). Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future. For this reason, institutions must equip students with the skills needed to use AI competently, in particular to perfect the art of _prompt engineering_, that is, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs. Universities will use AIs to improve learning and challenge the critical thinking of students. Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understand my own courses; automating the task is another way of cheapening the quality of the learning experience for students].

@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). On February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it. Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], that "[faculty] might not be able to stop the AI trend from growing", and so they might as well get on with the program and try to shape it as best they can.

@contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students". Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].] **MORE HYPE**

@@

A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. In the case of AI, critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making all sorts of claims to begin with^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed [flooding the zone with shit](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation). Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot possibly match [@maiberg__google_2024].].

A second discoursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.

In what follows, I do not plan to confront every single possible use case of AI as doing so plays into the intent of the Gish Gallop^[I have already spent close to 50 hours working on this essay, and it is still nowhere near an exhaustive apology or indictment of AI.], which is to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple other ways for teaching and learning to be excellent that do not require AI, and do not require the moral, environmental, social compromises demanded by AI[^Most discussions of AI as an everything tool acknowledge ].

@@

GenAIs are _deep learning models_ [@ibm_what_2021], so called because they are neural networks with _depth_, meaning many, many hidden layers. As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

@@

A regression model takes data as inputs, typically a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable). With these inputs, the model finds a conditional mean for the output $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables). The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).

Below is an example of the simplest regression model with only one independent variable (i.e., a bivariate model). The plot shows the pairs of $y$ and $x$ values, and the line is the model, the conditional mean $\hat{y}$. This model needs only two parameters: an intercept (the value of &\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$. Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
```{r simulate-simple-example}
set.seed(13253)

@@

```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 1. Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 2. Training data and model output compared"}

@@

Neural networks do something similar but using a net-like architecture that is more flexible at identifying the characteristics of the data. For instance, the following figure shows a relatively simple neural network trained using using the same data. This network is not "deep", as it has only one so called "hidden layer", comprised of two neurons between the input and the output layers. In comparison to the linear regression and its two parameters, this neural network uses seven parameters. Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the model contigent on the starting conditions.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 3. Simple Example of a Shallow Neural Network"}
knitr::include_graphics("images/nn1.png")
```

@@

```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 4. Training data and model outputs compared: linear regression and shallow neural network"}
ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

More complex neural network can be designed. The following plot is of a deeper network than the previous one, with two hidden layers, each consisting of five neurons. This model uses forty-six parameters. 
```{r train-deeper-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(242)
nn2 <- neuralnet(f, 
                data = df, 
                hidden = c(5, 5), 
                linear.output = T)
```

@@

```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 5. Example of a Deeper Neural Network"}
knitr::include_graphics("images/nn2.png")
```

@@

```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 6. Training data and model outputs compared: shallow neural network and deeper neural network"}
ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

Here are some key ideas to keep in mind:

- AIs are neural networks, which is to say, regression models.
- A regression model generates outputs that are regressions to the mean (i.e., statistically probable outputs).
- AIs are _very_ flexible non-linear regression models.
- Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters. OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be trillions.].
- AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
- The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
- Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think). The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train LLMs. The increasingly large datasets  essentially expanded the space where the model could interpolate.].

<!--

-->

**NOTE**: Labeling data is an essential, yet extremely expensive aspect of training GenAIs.

@@

Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it makes sense to use this technology. To begin, we must note the sheer number of examples of the use of AI by students. A report by @tyton_partners_generative_2023 that was sponsored by Turnitin found that 50% out of 1,600 post-secondary students who participated in the study, were users of GenAI (but only 22% of 1,000 faculty in the study). Use of AI by students is one of the reasons cited by some for universities to embrace AI [e.g., @power_3_2024]. McMaster's own internal research indicates that at least 61% of over 1,300 students polled, used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].

>>>>

Disclaimer: _All opinions expressed here are mine and any resemblance to those of my employer(s) is purely coincidental_. _No generative AI tool was used to research or write this essay, only good old-fashioned library and internet search skills_.

@@

- [Background](#background)  
- [The university's mission](#the-mission)  
- [What do AIs do?](#what-do-ais-do)  
- [A note on the terms of the conversation](#terms-of-debate)  
- [But really, what do AIs do?](#really-what-do-they-do)  
- [AI in teaching and learning](#ai-in-teaching-and-learning)  
- [What do GenAIs do for students?](#what-can-ais-do-for-students)  
- [What do GenAIs do to students?](#what-can-ais-do-to-students)  
- [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
- [What do GenAIs do to everyone else?](#what-do-ais-do-to-everyone-else)

@@

GenAI was hailed as nothing short of revolutionary. McMaster, for example, [stated](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:

@@

AI was quickly painted as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.

The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students.], has led to institutional responses that have been less than coherent. A case in point is Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].

@@

The sub-committees, in turn, convened panels to [provide recommendations](#ai-panel-tasks) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/). The [priority initiatives](#ai-panel-tasks) for the members of the AI Teaching and Learning Panel are to help shape the guidelines for teaching and learning at McMaster, a consequential task "for how educators take up generative AI in their course and assessment design and how students approach the use of this technology".

@@

Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption. This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn of the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023].] [@angwin_press_2024].

@@

(The last part of the statement above carries quite a bit of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.)

To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AIs, it is essential to understand what the technology claims to do that sets it apart from other tools, and what it in actuality does.

@@

Clearly, there is some tension implicit in the mission of a university. But being generous, we can take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with creativity, integrity, excellence, and quality.

@@

A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they have been presented quite deliberately as _everything tools_^[Almost as flexible as, well, a naturally intelligent human, but faster and supposedly cheaper.]. While hammers are understood to have a limited range of uses that rely on concentrating force, and word processors are mostly used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].

The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023].

Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023].

According to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024]. But there is no reason to stop at self-driving vehicles: McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023].

In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].

AI will even write code, perhaps helping itself to some self-programming along the way: AIs that write themselves, or AI self-help?

```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
knitr::include_graphics("images/AI-bug-fix.png")
```

The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].

Higher-education has not been spared the onslaught of promises.

@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.] for example, presents three innovative use cases for AI in higher education. First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".]. AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants. AI could also personalize the experience of consumers (also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.

Mr. Power's message to the higher-education sector? It is time for institutions to "level up": jump on the bandwagon, fellas, don't let it leave you behind!

In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school). Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future. For this reason, institutions must equip students with the skills needed to use AI competently, in particular train them to perfect the art of _prompt engineering_ (the ability to craft effective queries that prompt a GenAI to provide the most useful outputs). Universities will use AIs to improve learning and challenge the critical thinking of students. Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understanding my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].

@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). On February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it. Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.

Similarly hyperbolic claims aoput AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students". Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].]. A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "could", or "maybe will"):

- AI can provide personalised learning.  
- AI-powered systems can provide smart tutoring.  
- AI can automate grading.  
- AI can enable virtual classrooms.  
- AI can provide data insights for informed decisions.  
- Natural Language Processing (NLP) can enhance language learning.  
- AI can create interactive and adaptive learning materials.  
- AI can identify early signs of learning difficulties.  
- AI can aid in providing personalised professional development opportunities for teachers.
- AI can streamline administrative tasks.

Is it possible that AIs can really do all this?

@@

A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. To be sure, AIs _can_ do some things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation). Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].]. Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they do?

A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 65 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highligths the importance of institutions to provide the moral and ethical clarity that may elude individual instructors.].

@@

GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many hidden layers. As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

@@

A regression model takes data as inputs, typically a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable). With these inputs, the model finds a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables). The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).

Below is an example of the simplest regression model with only one independent variable (i.e., a bivariate model). The plot shows the pairs of $y$ and $x$ values, and the line is the model, the conditional mean $\hat{y}$. This model needs only two parameters: an intercept (the conditianl mean $\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$. Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
```{r simulate-simple-example}
set.seed(13253)

@@

```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 2. Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 3. Training data and model output compared"}

@@

Neural networks do something similar but using a net-like architecture that is more flexible at identifying the characteristics of the data. For instance, the following figure shows a relatively simple neural network trained using using the same data. This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers. In comparison to the linear regression and its two parameters, this neural network uses seven parameters. Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contigent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 4. Simple Example of a Shallow Neural Network"}
knitr::include_graphics("images/nn1.png")
```

@@

```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 5. Training data and model outputs compared: linear regression and shallow neural network"}
ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

More complex neural network can be designed. The following plot is of a network deeper than the previous one, with two hidden layers, each consisting of five neurons. This model uses forty-six parameters. 
```{r train-deeper-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(242)
nn2 <- neuralnet(f, 
                data = df, 
                hidden = c(5, 5), 
                linear.output = T)
```

@@

```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 6. Example of a Deeper Neural Network"}
knitr::include_graphics("images/nn2.png")
```

@@

```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 7. Training data and model outputs compared: shallow neural network and deeper neural network"}
ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

Here are some key ideas to keep in mind when thinking about AIs:

- AIs are neural networks, which is to say, regression models.
- A regression model generates outputs that are regressions to the mean (i.e., statistically probable outputs).
- AIs are _very_ flexible non-linear regression models.
- Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters. OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be in the trillions.].
- AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
- The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
- Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think). The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train AIs. The increasingly large datasets  essentially expanded the space where the model could interpolate.].

@@

Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it makes sense to use this technology. To begin, we must note the sheer number of examples of the use of AI by students. A report sponsored by Turnitin [@tyton_partners_generative_2023] found that 50% out of 1,600 post-secondary students in the study were users of GenAI (but only 22% of 1,000 faculty). Use of AI by students is a commonly cited reason for universities to embrace AI [e.g., @power_3_2024]. McMaster's own internal research indicates that at least 61% of over 1,300 students polled, used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].
